Supplementary material for "Hack Weeks as a model for Data Science Education and Collaboration"

### 2018-05-14 participants.tex

If merit-based selection is part of the evaluation process, organizers face the crucial decision whether to assess merit blindly or not. Because human decision makers tend to be swayed by unrelated characteristics including name~\cite{bertrand2004} or gender~\cite{mossracusin2012}, an initial merit selection blinded to demographic characteristics can be an effective way to counteract certain biases related to internal heuristics. A merit selection could then be performed via scores given independently by members of the organizing committee based on a set of pre-defined, explicit selection criteria. This type of blinded procedure tends to reduce biases during participant selection when committees would otherwise not consider diversity during their selection~\cite{bohnet2016}.

However, a blind selection purely on a figure of merit will be counterproductive if it excludes participants who might have had less exposure to certain technologies or fewer opportunities to learn certain skills: for example, if workshop organizers require a minimum level of programming experience, this could disadvantage candidates who have had fewer opportunities to learn programming due to structural inequalities. 
Additionally, blinding has been found to have negative effects on diversity for committees that have already had a strong commitment to diversity, because they often correct for structural inequalities by considering demographic variables during merit selection~\cite{behaghel2015unintended}.
In this case, it may be beneficial to construct selection criteria that explicitly consider diversity and inclusivity (as Neuro Hack Week has done; see also SM, Section 2.1). 
Because systemic biases likely also enter at the application stage (where underrepresented groups may be less likely to apply) organizers should consider oversampling traditionally disenfranchised groups compared to the population of applicants.

>>>>

If merit-based selection is part of the evaluation process, organizers face the crucial decision whether to assess merit blindly or not. Because human decision makers tend to be swayed by unrelated characteristics including name~\cite{bertrand2004} or gender~\cite{mossracusin2012}, an initial merit selection blinded to demographic characteristics can be an effective way to counteract certain biases related to internal heuristics. A merit selection could then be performed via scores given independently by members of the organizing committee based on a set of pre-defined, explicit selection criteria. This type of blinded procedure tends to reduce biases when committees would otherwise not consider diversity during their selection~\cite{bohnet2016}.

However, a blind selection purely on a figure of merit will be counterproductive if it excludes participants who might have had less exposure to certain technologies or fewer opportunities to learn certain skills: for example, requiring a minimum level of programming experience will likely disadvantage candidates who have had fewer opportunities to learn programming due to structural inequalities. 
Additionally, blinding has been found to have negative effects on diversity for committees that have already had a strong commitment to diversity, because they often correct for structural inequalities by considering demographic variables during merit selection~\cite{behaghel2015unintended}.
In this case, it may be beneficial to construct selection criteria that explicitly consider diversity and inclusivity (as Neuro Hack Week has done; see also SM, Section 2.1). 
Because systemic biases likely also enter at the application stage (where underrepresented groups may be less likely to apply) organizers should consider oversampling traditionally disenfranchised groups compared to the population of applicants.

### 2018-05-14 supplementary_material.tex

Since human decision makers also routinely overestimate their ability to forecast future performance of a candidate and underestimate intrinsic variance during selection procedures, ~\cite{highhouse2008} it may be beneficial to set a fairly tolerant threshold for acceptance based on merit criteria, and select the cohort from this meritorious pool via e.g.\ via an algorithm, imposing outside constraints on the selection based on the goals of the workshop.

@@

Some of our hack week tutorials follow the structure of the Software Carpentry model \cite{b:wilson-swc-lessons-2016}. The content is divided in to a series of lessons, each having a well-defined set of learning objectives, key questions, and expected outcomes. Interspersed throughout each lesson are a series of exercises which participants work through at their own pace. We frequently use a method drawn from Software Carpentry whereby students indicate their need for help on an exercise via posting a colored sticky-note on their computer. This provides a simple method to deliver help to those students who need it without interrupting the flow of the tutorial. We encourage instructors to develop online material to accompany their tutorials, either in the form of Jupyter Notebooks \cite{kluyver2016jupyter} or following the markdown templates provided by Software Carpentry. This way the hack week ``curriculum'' can be built up over time and students can refer to online content after the tutorial if they need to revisit certain topics. Depending on the content, some instructors convey concepts during the tutorial via ``live-coding'' while others display static code and their results. Either way, we find dividing code into short blocks, each with a specific task, helps participants from becoming saturated with information.
In some cases, tutorials require substantial software installation and data. Though we encourage participants to install these software dependencies on their own computers, so that they can continue using them after the tutorial, this is sometime so complicated that it slows down teaching. In some cases, we have experimented with \emph{Jupyterhub}\footnote{https://jupyterhub.readthedocs.io/en/latest/} to provide students with immediate access to a computational environment that provides teaching materials (e.g., Jupyter notebooks), software dependencies, and data \cite{holdgraf2017portable}. Providing access to an ephemeral consistent environment for the purpose of teaching has become even simpler with the recent advent of the \emph{Binder} system \cite{Holdgraf2017-pd, Titus_Brown_undated-pc}.

@@

The \textit{impostor phenomenon}, or \textit{impostor syndrome} (IS) is a dissonant feeling experienced by certain high-achieving individuals, that despite objective evidence to the contrary, they are in fact not as intelligent or capable as they appear.
Individuals with IS thus experience a fear of being ``found out'', shamed and expelled from their environment \cite{Clance1978-ef}.
Initial observations of the first Astro Hack Week conducted by data science ethnographer Brittany Fiore-Gartland\footnote{\url{http://astrohack week.org/blog/ethnographic-notes.html}} suggested that hack weeks are an environment prone to a particular kind of IS, namely that participants feel the need to be experts in multiple scientific and technical aspects of the activities pursued during a hack week.
This particular form of IS hinges to some degree on the design focus on diversity of backgrounds (everyone else seems to know something that you do not!) and might be further exacerbated by the expectation that attendants expose their ideas to public scrutiny, find collaborators in a very short amount of time, and not least produce and present a successful hack at the end of the week\footnote{see also this insightful blog post: \url{https://medium.com/astronomy-without-stars/the-horror-of-hack-days-52c6b52cfc3b}}.
The prevalence of IS at a hack week may be endemic to the format, and should thus be a major concern for any organizing committee.
This is because of the chilling effect it tends to have on participants and the community as a whole, and particularly on women (in particular in fields in which women are under-represented) and members of ethnic and racial minorities, correlating with anxiety and other forms of mental distress \cite{Parkman2016-ro}.
Less severely, another major concern is that IS inhibits risk taking: participants experiencing it will be less likely to ask a question, to put forward an idea for a hack, to be pro-active about forming new collaborations.
 Many of the goals of a hack week, including the successful completion of projects, lateral knowledge transfer, as well as community building are hampered.
We are working within our hack weeks to mitigate IS using various techniques.
With respect to minority participants, ensuring adequate representation can decrease feelings of otherness and may help reduce IS.
More generally, being open about the presence and prevalence of IS can help participants feel more at ease. %(e.g. one Astro Hack Week participant remarked in a survey, \textit{''I really appreciated the direct acknowledgement of impostor syndrome on the first day.
%I think it helped ease the feeling!!``}).
Additionally, role models are very effective at encouraging positive behaviour: asking participants with prior experience at hack weeks at all academic levels to ask questions, even when they might know the answer, can help foster an inclusive environment that rewards risk taking. We also invited our organizing committees to model open communication about their own handling of IS through the course of their career. We found that transparency about one's own uncertainty, espcially from community leaders, helped create a hospitable atmosphere and removed barriers.

@@

\begin{figure*}[h!]
%\begin{center}
%\begin{subfigure}[t]{\textwidth}
\centering
%\caption{}
\includegraphics[width=\textwidth]{Q19_5_Q27_stackedbars.eps}
%\end{subfigure}
\caption{}
\label{fig:survey}
%\end{center}
\end{figure*}

>>>>

Since human decision makers also routinely overestimate their ability to forecast future performance of a candidate and underestimate intrinsic variance during selection procedures~\citep{highhouse2008}, it may be beneficial to set a fairly tolerant threshold for acceptance based on merit criteria, and select the cohort from this meritorious pool via e.g.\ via an algorithm, imposing outside constraints on the selection based on the goals of the workshop.

@@

Some of our hack week tutorials follow the structure of the Software Carpentry model \citep{b:wilson-swc-lessons-2016}. The content is divided in to a series of lessons, each having a well-defined set of learning objectives, key questions, and expected outcomes. Interspersed throughout each lesson are a series of exercises which participants work through at their own pace. We frequently use a method drawn from Software Carpentry whereby students indicate their need for help on an exercise via posting a colored sticky-note on their computer. This provides a simple method to deliver help to those students who need it without interrupting the flow of the tutorial. We encourage instructors to develop online material to accompany their tutorials, either in the form of Jupyter Notebooks \citep{kluyver2016jupyter} or following the markdown templates provided by Software Carpentry. This way the hack week ``curriculum'' can be built up over time and students can refer to online content after the tutorial if they need to revisit certain topics. Depending on the content, some instructors convey concepts during the tutorial via ``live-coding'' while others display static code and their results. Either way, we find dividing code into short blocks, each with a specific task, helps participants from becoming saturated with information.
In some cases, tutorials require substantial software installation and data. Though we encourage participants to install these software dependencies on their own computers, so that they can continue using them after the tutorial, this is sometime so complicated that it slows down teaching. In some cases, we have experimented with \emph{Jupyterhub}\footnote{https://jupyterhub.readthedocs.io/en/latest/} to provide students with immediate access to a computational environment that provides teaching materials (e.g., Jupyter notebooks), software dependencies, and data \citep{holdgraf2017portable}. Providing access to an ephemeral consistent environment for the purpose of teaching has become even simpler with the recent advent of the \emph{Binder} system \citep{Holdgraf2017-pd, Titus_Brown_undated-pc}.

@@

The \textit{impostor phenomenon}, or \textit{impostor syndrome} (IS) is a dissonant feeling experienced by certain high-achieving individuals, that despite objective evidence to the contrary, they are in fact not as intelligent or capable as they appear.
Individuals with IS thus experience a fear of being ``found out'', shamed and expelled from their environment \citep{Clance1978-ef}.
Initial observations of the first Astro Hack Week conducted by data science ethnographer Brittany Fiore-Gartland\footnote{\url{http://astrohack week.org/blog/ethnographic-notes.html}} suggested that hack weeks are an environment prone to a particular kind of IS, namely that participants feel the need to be experts in multiple scientific and technical aspects of the activities pursued during a hack week.
This particular form of IS hinges to some degree on the design focus on diversity of backgrounds (everyone else seems to know something that you do not!) and might be further exacerbated by the expectation that attendants expose their ideas to public scrutiny, find collaborators in a very short amount of time, and not least produce and present a successful hack at the end of the week\footnote{see also this insightful blog post: \url{https://medium.com/astronomy-without-stars/the-horror-of-hack-days-52c6b52cfc3b}}.
The prevalence of IS at a hack week may be endemic to the format, and should thus be a major concern for any organizing committee.
This is because of the chilling effect it tends to have on participants and the community as a whole, and particularly on women (in particular in fields in which women are under-represented) and members of ethnic and racial minorities, correlating with anxiety and other forms of mental distress \citep{Parkman2016-ro}.
Less severely, another major concern is that IS inhibits risk taking: participants experiencing it will be less likely to ask a question, to put forward an idea for a hack, to be pro-active about forming new collaborations.
 Many of the goals of a hack week, including the successful completion of projects, lateral knowledge transfer, as well as community building are hampered.
We are working within our hack weeks to mitigate IS using various techniques.
With respect to minority participants, ensuring adequate representation can decrease feelings of otherness and may help reduce IS.
More generally, being open about the presence and prevalence of IS can help participants feel more at ease. %(e.g. one Astro Hack Week participant remarked in a survey, \textit{''I really appreciated the direct acknowledgement of impostor syndrome on the first day.
%I think it helped ease the feeling!!``}).
Additionally, role models are very effective at encouraging positive behaviour: asking participants with prior experience at hack weeks at all academic levels to ask questions, even when they might know the answer, can help foster an inclusive environment that rewards risk taking. We also invited our organizing committees to model open communication about their own handling of IS through the course of their career. We found that transparency about one's own uncertainty, espcially from community leaders, helped create a hospitable atmosphere and removed barriers.

@@

\begin{figure*}[h!]
%\begin{center}
%\begin{subfigure}[t]{\textwidth}
\centering
%\caption{}
\includegraphics[width=\textwidth]{Q19_5_Q27_stackedbars.eps}
%\end{subfigure}
\caption{}
\label{fig:survey}utbt
%\end{center}
\end{figure*}
